The Forecasting Trap -- notes

### 2022-02-15 manuscript/round3-manuscript.Rmd
:pen:

>>>>

First, I will introduce a motivating example in which we will consider two reasonable process-based models, A and B. 
Model B will produce very accurate forecasts, but lead to much worse outcomes than Model A.
These accurate forecasts in Model B are not the result of chance or of over-fitting the data.

# A motivating example

To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.

I will use the term "model" to refer to any set of equations or code that can be used to produce a forecast.
This term thus includes not only traditional process-based models, but could also includes statistical forecasting methods (e.g. ARIMA, @arima), non-parametric approaches such as empirical dynamical modeling [@Ye2015], or machine learning. 
Most such models must first be calibrated to historical data before they can produce a forecast, e.g. by parameter fitting, expert knowledge, or some other means. 
Because different choices for those parameters create different forecasts, I will refer to those different parameterizations as different models.
It is of course possible for a decision-maker to consider forecasts coming from multiple structurally different models simultaneously, and potentially assigning different weights to each model.
As more data becomes available, it is possible to update model parameters or the weights assigned to models.
I will examine such approaches for model ensembles and model updating further on.

In these examples, we will focus on situations in which our 'data' comes from a model simulation rather than empirical sources.
This enables us to conduct experiments that would be often impossible or unethical to perform in the real world: for instance, does a given fishery experience better long-term outcomes when managed according to forecasts derived from model 1 or from model 2?
The ability to perform many replicate simulations will also allow us to avoid results that could be due to chance alone.

We will base this data-generating simulation on the five-species model of @Brias2021.
In this model, three competing species of river herring are preyed upon by by striped bass and double-crested cormorants.
Managers face the multiple competing objectives of promoting economically valuable fisheries for striped bass and herring, while the double-crested cormorants are a target of conservation.
Following the formulation of @Brias2021, let us assume managers seek to maximize a weighted sum of composed of the total number of cormorants and the value of harvest from both herring fishery, and the striped bass fishery.
We will assume the manager has decided to balance conservation and economic objectives, with 50% weight allocated to cormorant population and 25% to each fishery.
We will assume that the manager must select a fixed fishing effort $u_{\textrm prey}$ and $u_{\textrm predator}$ that determines the annual mortality of herring and bass respectively. 
In practice, managers may have flexibility to adjust fishing effort annually, increasing the complexity of the decision involved.
We will explore more complex decision-processes later, but note that in many ecological settings, choices available to managers are highly constrained and frequent adjustment of policies is costly or impossible [@ @ @].

Reality is always more complex than even our richest models [@Schindler2015].
In the real world, we can never have the 'true model' among our candidates -- after all, models seek to represent only the most salient features for their purpose [@Getz2018].
To reflect this gap between model and real-world complexity, we assume our manager does not know the underlying process (Eq 1), but must choose between two alternative three-dimensional models.
Both models will treat the herring as if it were a single species, rather than three distinct species.
This lumping of data is typical of one of the many assumptions any model must make, and can be justified in terms of ecological similarities and/or data limitations. 
A second important aspect of real-world systems and realistic models is that not all states can be measured perfectly. 
To reflect this, we will assume the manager has only estimates of the bass and cormorant populations.

Beyond this, each of our alternative models makes somewhat different assumptions about the biology, as seen in comparing the dynamical equations for Model A to Model B.  
Model A does not explicitly capture reflect the role of predation on the herring population, and oversimplifies the dynamics of the cormorant population as tracking at a fixed fraction of the herring population.  
These assumptions are not strictly true, but are justifiable in certain limits -- modeling is the science of approximation.
In contrast, under Model B, the trophic structure more closely reflects the true model, with the herring population growth rate decreasing when either cormorant or bass populations increase.

Model parameters are selected for each model consistent with limited 'historical' data from stochastic simulation.
Given either forecast model, it straight forward to determine the harvest policy  $\lbrace u_{\textrm prey}, u_{\textrm predator} \rbrace$ which maximizes the stated management objective. 
We perform a standard Nelder-Meade optimization routine on each forecast to determine the optimal policy.
We also note the forecasts each model predicts it will see using policy $T_\{max}$ years into the future.

We can simulate what the actual management outcomes and observed species dynamics are under the policy derived from each forecast. 
We compare how well each model did in predicting the resulting population dynamics across each species (considering the herring population as the sum across all three herring species), as well as the realized utility defined by the management objective.
Because we are using a simulation, we can also go one step further and compare the policy and mean net utility to that of the optimal strategy for the true model.

### 2022-02-15 notebook/alternate-models.Rmd
:pen:

@@

@@

```{r}
model_B <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      R1 =  1.15, #1.0213,
                      A11 = 1,
                      A14 = 0.1, # 0.7252,
                      A15 = 0.1, # 0.2445,
                      A41 =  0.8, #0.7252,
                      A51 = 0.3, # 0.2445,
                    reps = 40) {

B <- optim(c(0.5,0.1), u_B)
B$value
model_B_policy <- clip(B$par, 0, 1)
model_B_policy

```{r}
## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
Tmax <- 5
x <- c(0,0.58)

obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

  #scoringRules::crps_sample(predict$x1, )

}

>>>>

The model which gives the best forecast does not always lead to the best decision.

@@

```{r}
model_B <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      R1 =  3, #1.15, #1.0213,
                      A11 = 8,
                      A14 = 0.9, # 0.7252,
                      A15 = 0.9, # 0.2445,
                      A41 =  0.1, #0.7252,
                      A51 = 0.3, # 0.2445,
                    reps = 40) {

B <- optim(c(.2,0.01), u_B)
model_B_policy <- clip(B$par, 0, 1)
model_B_policy

sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax, sigma= 0.05)
obs <- sim %>% format_sim()
predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

reward(df = sim, x = x) / 40
```

```{r}
obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")

predict_stats_B <- predict_B %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted_B")

bind_rows(obs_stats, predict_stats_B) %>% 
  filter(species != "herring") %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_point(aes(shape=type)) +
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type))

```{r}
## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
Tmax <- 10
x <- c(0,0.58)

sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax, sigma=0.05)
obs <- sim %>% format_sim()
predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

reward(df = sim, x = x) / 40
```

```{r}
obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")

predict_stats <- predict %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted_A")

bind_rows(obs_stats, predict_stats) %>% 
  filter(species != "herring") %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_point(aes(shape=type)) +
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type))
```

}
```

## unfished

```{r}
## True observations vs forecasts, 5 steps ahead, unfished, starting with default initial condition
Tmax <- 20
x <- c(0,0.)

obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], sigma=0.05, Tmax = Tmax) %>% format_sim()
predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

Tmax <- 5
x <- c(0,0)
x <- model_B_policy
#x <- c(0.,0.58)  # Following the optimal policy under model_B

@@
